Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-04-27 demystify.tex
updates

The focus of adcom is to evaluate your background and interest in research to see if you \textbf{fit} into our program and can handle a PhD in CS, which is a \emph{research} degree!  To evaluate your profile, we consider the following key indicators.

@@

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining admission to top CS PhD programs. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), these reference letters are generally effective and have helped many students get into top PhD programs in the US.

@@

In your SOP, focus on research motivation and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidence. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on project X, then show some results (paper submitted (or even rejected), achieved X\% performance improvement over ..). Also, do not use more than two pages (references can be on the third page). \href{https://cs-sop.org/}{Many SOP samples} are available.

@@

In short, you \emph{do not need} to formally taking CS courses, you just need to show that you have these essential knowledge, e.g., through the mentioned ways. Many universities are well-aware that incoming graduate students might not have all the technical background, so they often have a \emph{``bridge''} courses to help students catch up.  For example, GMU has four bridge courses corresponding to the four core areas above that incoming students can take to catch up on their CS knowledge.

@@

If you have an MS then some course work \emph{might be} transferred for course credits, which saves a bit of time. But overall don't count on this, especially if your MS is not from the US.

@@

The first 2 years you typically take coursework (somewhat equivalent to an MS study), find an adviser, and learn how to do research.  The next 2--3 years you focus on your research, form a dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for a job.
In many cases you might take a summer or two off to do an internship to get additional research opportunities.
The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

>>>>

The focus of adcom is to evaluate your background and interest in research to see if you \textbf{fit into our PhD program}! The emphasis here is \emph{fitting}, which varies from school to school, faculty to faculty, and even from year to year.  However, in general, to evaluate your profile, we consider the following key indicators.

@@

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF2.0 program}, which has helped many students in gaining admission to top CS PhD programs. VEF2.0 invites US faculty members from leading institutions to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily just the interview), these reference letters are generally effective and have helped many students get into top PhD programs in the US.

@@

In your SOP, focus on research motivation and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidence. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on project X, then show some results (paper submitted (or even rejected), achieved X\% performance improvement over ..). Also, do not use more than two pages (references can be on the third page). Many SOP samples are \href{https://cs-sop.org/}{available here}.

@@

In short, you \emph{do not need} to formally taking CS courses, you just need to show that you have this essential knowledge, e.g., through the mentioned ways. Many universities are well-aware that incoming graduate students might not have all the technical background, so they often have a \emph{``bridge''} courses to help students catch up.  For example, GMU has four bridge courses corresponding to the four core areas above that incoming students can take to catch up on their CS knowledge.

@@

If you have an MS then some course work \emph{might be} transferred for course credits, which saves a bit of time. But don't count on this, especially if your MS is not from the US.

\begin{commentbox}[Vu:]
    I start my PhD with an MS in CS from a US university.  I found that the MS helped me a lot in research, but not with course work and I had to retake many courses.  In general, don't worry if you don't have an MS. But also don't feel that you wasted your time if you have an MS, as it can help you in research.
  \end{commentbox}

@@

The first two years you typically take coursework (somewhat equivalent to an MS study), find an adviser, and learn how to do research.  The next 2--3 years you focus on your research, form a dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for a job.
In many cases you might take a summer or two off to do an internship to get additional research opportunities.
The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

\begin{commentbox}[Vu:]
    I start my PhD with an MS, and it took me 7 years (Fall'2007-- Fall'14). I did spend half a year doing internship at Naval Research Lab and 1 year doing internship at Google
  \end{commentbox}
